---
name: Thomaz F. S. Bastiaanssen
date: "`r format(Sys.time(), '%B, %Y')`"
qualifications: BSc, MSc, PhD
address: Cork, Republic of Ireland
#profilepic: pic.jpg
email: thomazbastiaanssen@gmail.com
github: thomazbastiaanssen
linkedin: thomaz-bastiaanssen-753242b0
twitter: thomazsan
aboutme: Bioinformatics, Microbiome-Gut-Brain Axis, Theoretical Ecology
headcolor: 414141
output: vitae::moderncv

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(vitae)
```

# Professional Summary

I am a bioinformatician interested in the interplay between the gut microbiome and host mood and mental health. 
The focus of my research currently lies in understanding microbiome-gut-brain communication from a theoretical ecology/bioinformatics perspective. In particular, I am interested in understanding the role of stability and volatility of the gut microbiome in anxiety and depression and in integrating different types of ‘omics data in a biologically interpretable manner. As a bioinformatician, I value clear and easily interpretable analysis of complex data in order to promote interdisciplinary collaboration.

# Employment
```{r Employment}

#use this command to edit your r profile in a new window
#usethis::edit_r_profile()

#enter this line followed by an empty new line. 
#ORCID_TOKEN="46ec8f60-5d1d-4886-b5ba-3d7bd9ec8bf8"
           
orcid_emp_data <- do.call("rbind",
  rorcid::orcid_employments("0000-0001-6891-734X")$`0000-0001-6891-734X`$`affiliation-group`$summaries
)

notes = c(`Postdoctoral Researcher` = "Lead bioinformatician for the Cryanlab")

orcid_emp_data %>%
  cbind(notes) %>%
  detailed_entries(
    what = `employment-summary.role-title`,
    when = glue::glue("{`employment-summary.start-date.year.value`} - now"),
    with = `employment-summary.organization.name`,
    where = `employment-summary.organization.address.city`, 
    why = notes
  )

```

# Education
```{r education}

#use this command to edit your r profile in a new window
#usethis::edit_r_profile()

#enter this line followed by an empty new line. 
#ORCID_TOKEN="46ec8f60-5d1d-4886-b5ba-3d7bd9ec8bf8"
           
orcid_data <- do.call("rbind",
  rorcid::orcid_educations("0000-0001-6891-734X")$`0000-0001-6891-734X`$`affiliation-group`$summaries
)

notes = c(PhD = "Thesis Title: Mining the Microbiome for Markers of Microbiota-Gut-Brain Communication and Mental Health", 
          MSc = "Research Trajectory: Molecular and Cellular Life Sciences - Bioinformatics", 
          BSc = "Majored in Biology with a Minor in Art History")

orcid_data %>%
  cbind(notes) %>%
  detailed_entries(
    what = `education-summary.role-title`,
    when = glue::glue("{`education-summary.start-date.year.value`} - {`education-summary.end-date.year.value`}"),
    with = `education-summary.organization.name`,
    where = `education-summary.organization.address.city`, 
    why = notes
  )

```

# Research Skills

```{r skills,results='asis', width = 40000}

skills <- read_csv("skills.csv")

skills_formatted <- skills %>% 
  #mutate(skill = if_else(is.na(details), glue::glue("{skill}"), glue::glue("{skill})"))) %>% 
  group_by(group) %>% 
  summarize(skills = glue::glue_collapse(skill, sep = " $\\bullet$ "))

for (group in skills_formatted$group) {
  cat(
    glue::glue("\\textbf{<<tools::toTitleCase(group)>>}:", .open = "<<", .close = ">>"),
    dplyr::filter(skills_formatted, group == !!group)$skills,
    "\n\n\n"
  )
}
#pubs$authors = temp
```
\newpage

# Publications

```{r publications}

pubs = scholar::get_publications("QktcRXUAAAAJ") %>% 
  dplyr::mutate(author = author %>% 
                  as.character %>% 
                  stringr::str_trim(),
                journal = journal %>% 
                  replace(journal %in% "bioRxiv", "BioRxiv"), 
                first_author = case_when(stringr::str_starts(author, "TFS Bastiaanssen") ~ TRUE,
                                         TRUE ~ FALSE),
                preprint = case_when(journal %in% c("bioRxiv", "BioRxiv") ~ TRUE,
                                         TRUE ~ FALSE)) %>% 
  dplyr::arrange(desc(year))

#Account for shared firsts
pubs[pubs$title == "Microbiota from young mice counteracts selective age-associated behavioral deficits",]$first_author <- TRUE

#pubs$authors = temp
```


**Selected first author**

```{r}
pubs %>% 
  dplyr::filter(first_author, !preprint) %>% 
  dplyr::filter(cites > 10) %>%
  detailed_entries(
    what = title,
    when = as.character(year),
    with = author,
    where = journal, 
    .protect = TRUE
  )
```

**Selected non-first author**

```{r}
pubs %>% 
  dplyr::filter(!first_author, !preprint) %>% 
  dplyr::filter(cites > sort(cites, decreasing = T)[5]) %>%
  dplyr::arrange(-cites) %>%
  detailed_entries(
    what = title,
    when = as.character(year),
    with = author,
    where = journal, 
    .protect = TRUE
  )
```


<!-- # Relevant Skills -->
<!-- ```{r} -->
<!-- cvskill("Data Analysis", c("Visualisation", "Data Cleaning", "Forecasting", "Modelling", "Communication", "Inference", "Machine Learning", "Web Scraping")) -->
<!-- cvskill("Programming", c("R", "Python", "SQL", "Java", "C", "C++", "HTML/CSS", "TeX", "MATLAB")) -->
<!-- ``` -->
